# R Packages Notes

## purrr

[purrr package](https://purrr.tidyverse.org/). "This package completes R's functional programming tools with missing features present in other programming languages."

The guide to purrr on the tidyverse site is limited but it recommends the [iteration chapter](https://r4ds.had.co.nz/iteration.html) in R for data science. There are detailed [cheat sheets](https://github.com/rstudio/cheatsheets/blob/master/purrr.pdf) which also include use of nested data frames with tidyr functions that often link with purrr functions. Also see,

+ [Jenny Bryan purrr tutorial](https://jennybc.github.io/purrr-tutorial/index.html). 
+ An example-based [purrr intro](https://emoriebeck.github.io/R-tutorials/purrr/).
+ [Purrr cookbook](http://colinfay.me/purrr-cookbook/)

Also related to this are more general row wise operations on dataframes. Jenny Bryan gave an [Rstudio seminar on row oriented workflows](https://github.com/jennybc/row-oriented-workflows). This gives of options and things to consider but doesn't go into specific detail on the methods. More recently (as of 03/2020), there is a [rowwise operations vignette](https://dplyr.tidyverse.org/dev/articles/rowwise.html) for dplyr which is based on the most recent (beta) developments using `across()`. It uses `rowwise()` extensively (not mentioned in Jenny Bryan's talk) and says it can avoid use of purrr. It says, `rowwise()` was questioning for sometime, with purrr being preferred, but they now regard `rowwise()` as being easier to use and it is no longer questioning. This is the most up-to-date methodology for dplyr and dataframes but purrr is used more generally.

Brief notes from the R4DS chapter:

+ `map()` has a `.f` argument which can be a function, formula, character vector or integer vector. Use `~` to create an anonymous function. In an anonymous function `.` refers to current list element.
+ `map()` is identical to `lapply()` but adds in some shortcuts.
+ Shortcuts extract named elements or by position e.g. `map_dbl("r.squared")` or `map(2)`.
+ The `map_*()` variants are like `vapply()` but need less typing. `vapply()` can ouput matrices while `map_*()` can't.
+ `map2()` and `pmap()` use multiple related inputs. `invoke_map()` applies multiple functions. `walk()` returns side_effects. See [https://purrr.tidyverse.org/reference/map2.html](https://purrr.tidyverse.org/reference/map2.html). 
+ There are lots of other useful functions in the package e.g. `safely()`, `possibly()`, and `quietly()` manage errors in parallel actions. 

Example `pmap()` use with a df:

```{r}
ari <- function(sum, a, b) {
  if (sum){
    return(a + b)
  }
  a * b
}

tibble(sum = c(T, F, T, F), a = 1:4, b = 5:8) %>% 
  mutate(out = pmap_int(., ari))
```

Use `...` in helper function to absorb extra cols in df/list passed to `pmap()`.

## dplyr

**THESE NOTES ARE OLD**

`count()` and `tally()` can be used to count unique values. The following are equivalent:

```{r eval=F}
mtcars %>% group_by(cyl) %>% summarise(n=n())
mtcars %>% group_by(cyl) %>% tally
count(mtcars, cyl)
```

By using the weight argument to `count()` it can also be used to sum a column, so these two are equivalent:

```{r eval=F}
mtcars %>% group_by(cyl) %>% summarise(tot = sum(mpg))
count(mtcars, cyl, wt=mpg)
```

Note: there are slight differences if working with non-tibble data since `count()` does not convert to a tibble while `group_by()` does.

### `mutate()` and `Summarise()`

Grouping and summarise creates a new tibble with new variables which you name in `summarise`. Therefore summarise and mutate are related and have similar structure. The difference is that `summarise` uses **summary or aggregation functions** which take a vector of values and return a single value while `mutate` uses **window functions** which take a vector of values and return a vector of the same length. Therefore `mutate` won't work directly with a user defined function that has scalar inputs. A way around this is to use `rowwise()` before the mutate call. See [this discussion](https://stackoverflow.com/questions/21818181/applying-a-function-to-every-row-of-a-table-using-dplyr). It is probably better to write a vectorised version of the function I wish to use.

## Selecting columns as inputs to `mutate()`

This is where I want to use mutate where the new variable is a function of many columns such that I don't want to type them all out explicitly. Note that this is different from selecting columns _to_ mutate. I found several Stack Overflow threads wanting to do this for a sum for which `mutate(sum = rowSums(.[1:4]))` or `mutate(sum=Reduce("+",.[1:4]))` could be used (possibly with `na.rm=T` argument to `rowSSums()`). The indices in square brackets could be replaced by a character vector of variable names also. I'm not as sure about the more general case asked [here](https://stackoverflow.com/questions/28095526/summarise-over-all-columns). `Reduce()` can be used more generally and might be enough (I think it only works with pairwise operations). `?Reduce` gives some other possibly useful functions too. Remember `rowwise` is needed for non-vectorised functions. The link gives an answer using `do()` and `unlist()` which might help understanding.

## Other notes

[pivot_](https://speakerdeck.com/yutannihilation/a-graphical-introduction-to-tidyrs-pivot-star). Nice slides on how to use `pivot_longer()` and `pivot_wider()`.
