# Cookbook for R

This is a place to put miscellaneous solutions to routine problems I have met in R. 

## Data frame rows to a list

Data frames naturally split into lists by columns. I wanted to make a list where each element holds a single row of the data frame as a list with elements named with column names. The base option works fine, but I later found `purrr::transpose()`. The `group_split()` function is currently marked as _experimental_.

```{r, eval = FALSE}
split(dt, seq(nrow(dt))) %>% map(as.list)
rowwise(dt) %>% group_split %>% map(as.list)
purrr::transpose(dt)
```

The first options adds numeric names at the top level, while the others leave it unnamed. `setNames` or `rlang::set_names()` can be used to name or rename the list.

Much of this came from [thread](https://stackoverflow.com/questions/3492379/data-frame-rows-to-a-list).

## Passing list of named parameters to a function

Arguments can be passed to another function inside a function easily using `...` but the issue here was that the arguments to the first function were a named list e.g. `list_of_args = list(a = 1, b = 3)` where I wanted to call `f(list_of_args)` and have the inner function handle this e.g. `f2(...)`.

Firstly the contents of `...` can be captured in the function using `c(...)` or `list(...)`. Secondly, arguments in a list can be passed directly to a function using `do.call(fun_name, args_list)`. See this [thread](https://stackoverflow.com/questions/9129673/passing-list-of-named-parameters-to-function) (mainly about managing lists). Interesting how the output of `c()` depends on the deepest list level in its inputs. 

Maybe useful in this context is `alist()` which handles its arguments as if they describe function arguments.

## Shiny - Dynamic `selectInput()`

I used the following code to get a drop-down menu that changed according to other dynamic inputs.

```{r eval = FALSE}
output$x_options <- renderUI({
  selectInput("x_var", "Variables to show on x-axis:", input$clust_vars)
})
output$y_options <- renderUI({
  selectInput("y_var", "Variables to show on y-axis:", input$clust_vars)
})

inputPanel(
  uiOutput("x_options"),
  uiOutput("y_options"),
  sliderInput("label_n", "Number to label from each cluster:",
              min = 1, max = 5, value = 2, step = 1)
)  
```

## Remove NAs from Data Frame

To return a data frame without rows with `NA`s:

+ `na.omit()` - omits any column with a missing value. data.frame has a faster version.
+ `complete.cases()` - returns a logical vector of which rows don't contain any `NA`s. Can use this to selectively exclude columns e.g. `dat[complete.cases(dat[ , 5:6]),]` will just check cols 5 and 6 for `NA`s.
+ `tidyr::drop_na()`. Optional tidy-select arguments to only inspect some columns.
