# Tidy Evaluation

```{r tidyeval_setup, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = TRUE, eval = TRUE)
library(tidyverse)
```

Tidy evaluation is a framework for metaprogramming in R. The notes here will focus on that framework but will also cover broader ideas such as standard and non-standard evaluation (NSE). There has been a lot of movement on this in the past year or two so it may change further, but these notes aim be current as of writing (April 2020).

## Links

A difficulty in this area is that it is hard to tell what is current since the area is still evolving. I also have some confusion over terminology and the packages involved. I think the core of tidy eval is contained in rlang, while tidyselect adds select helpers and other functions on top. 

+ [Programming with dplyr](https://dplyr.tidyverse.org/articles/programming.html). As of dplyr 1.0, I take this as the latest word on how tidy evaluation should be used in dplyr.
+ [rlang 0.4.0 blog post](https://www.tidyverse.org/blog/2019/06/rlang-0-4-0/). Introduces `{{` operator. This isn't mentioned in the Tidy Evaluation bookdown. It appears to be a shorthand for some of the methods given there.
+ [rlang blog 2020-02](https://www.tidyverse.org/blog/2020/02/glue-strings-and-tidy-eval/). Brings glue strings into tidy eval. Used together with `{{`.
+ [Data pronouns for tidy evaluation vignette](https://rlang.r-lib.org/reference/tidyeval-data.html). This gives `.data` and `.env` pronouns to be explicit in data masking. This might be what I should be doing in my packages.
+ [tidyselect package](https://tidyselect.r-lib.org/index.html) imports rlang (and other packages). Listed as dependency for dplyr but its functions are not loaded with the tidyverse. There are two vignettes: [Implementing tidyselect interfaces](https://tidyselect.r-lib.org/articles/tidyselect.html) and [Technical description of tidyselect](https://tidyselect.r-lib.org/articles/syntax.html). 
+ [Tidy Evaluation book](https://tidyeval.tidyverse.org/). Now superseded. Instead they link to the Programming with dplyr vignette and [using ggplot in packages](https://ggplot2.tidyverse.org/articles/ggplot2-in-packages.html).
+ The [metaprogramming chapter in Advanced R](https://adv-r.hadley.nz/metaprogramming.html) has more on the general area.

The documentation of the various functions is useful too and my notes will include these. This [post on background of quotation](https://milesmcbain.xyz/the-roots-of-quotation/amp/) might be useful for deeper understanding. 

Also see the [Tidyverse Design Guide (book)](https://design.tidyverse.org/).

## Tidy Evaluation

Most dplyr functions use non-standard evaluation (NSE) which means they capture the expression you type and evaluate it in a custom way. This creates ambiguity depending on where variables were defined and can make it hard when passing column names via function argument. 

Most dplyr arguments are not _referentially transparent_. That means you can’t replace a value with a seemingly equivalent object that you’ve defined elsewhere.

**Quoting** is the action of capturing an expression instead of evaluating it. All expression-based functions quote their arguments and get the R code as an expression rather than the result of evaluating that code. Examples are *formulas* and `quote()`, which returns an expression. Expressions formed by `quote()` or `~` have tree structures. The double quote `"` is not a quoting operator since it generates a string, not an expression. A formula captures the execution environment as well as the code. 

When an object keeps track of an environment, it is said to have an enclosure. This is the reason that functions in R are sometimes referred to as closures. A one-sided formulas is given a special name: **quosures**. One-sided formulas are quotes (they carry an expression) with an environment.

**Quasiquotation** allows you to evaluate directly inside an expression that is otherwise quoted. All expression-based functions in the tidyeval framework support quasiquotation. Unquoting cancels quotation of parts of an expression. There are three types of unquoting:

+ Basic. Functional form is `UQ()` and the syntactic shortcut is `!!`.
+ Unquote splicing. Functional form is `UQS()` and the syntactic shortcut is `!!!`. It takes a vector and inserts each element of the vector in the surrounding function call.
+ Unquoting names. Uses the `:=` operator.

Note that `UQ()` and `UQS()` are [deprecated now](https://rlang.r-lib.org/reference/nse-force.html). An alternative explanation of these terms from the [roots of quotation post](https://milesmcbain.xyz/the-roots-of-quotation/amp/) explains this better:

+ *quote* transforms code into data.
  + That data being the machine-readable symbolic structure needed to *formally* evaluate it.
+ *quasiquotation* is a mode that assumes we want all code transformed to data.
+ *unquote* flags code that we wish to have the result of transformed to data.
  + In the context of quasiquotation.

So within dplyr functions the default is to use NSE and quote arguments (quasiquotation is used). Then unquoting is needed to force evaluation of any parts we want evaluated.

The three pillars of tidy evaluation are quasiquotation, quosures, and the data mask.

Functions (or more specifically, their arguments) can either be quoting or evaluating. Evaluating is the standard, while quoting captures the argument to process in a custom way. `library()` is a quoting function so it's first argument does not have to be a string or exist as an object. `[[` is an evaluating function while `$` is a quoting function. See [this chapter](https://tidyeval.tidyverse.org/sec-up-to-speed.html) for more details.

### Tidyeval Functions

I have tried to match-up base functions with tidyeval equivalents to understand where the newer functions come from.

+ Base R `quote()` "simply returns its argument". Similar to `rlang::expr()`. These return an expression without evaluating it (they are *defusing* operators). I don't think they record the environment. 
+ Formulas use the `~` operator. I am unsure how they differ from `quote()`. More specific structure?
+ `rlang::exprs()` is similar to `alist()`.
+ `rlang::quo()` is another defusing operator, but which returns an expression with an environment attribute (a *quosure*). It has similarities with `expr()`. It is possible it is replacing it - I am not sure they are both mentioned in the same document. 
+ `rlang::enquo()` is a user facing version of `quo` designed to deal with function arguments (`quo()` is developer facing). The base equivalent is `substitute()`.
+ I'm not sure about `sym()` and `syms()` (there is also `ensym()` and `ensyms()`. They convert strings to symbols. A defused expression can be a *call* or a *symbol*. Symbols represent named objects. I think they are stricter versions of the other quoting functions that restrict user input to strings.
+ The `!!` and `!!!` operators are unquoting (or forcing) operators which force evaluation of quoted code. There are function call versions of these, `UQ()` and `UQS()`, but Advanced R says these aren't really function calls, so they may be discouraged.
+ The `{{` operator is a recent development. It provides a shorthand for quote-unquote operation e.g. `!!enquo(x)`.
+ I used to use the base `as.name()` function with `!!`. I don't know how this relates, maybe similar to `sym()` (the base `as.symbol()` is identical to `as.name()`).
+ `eval_tidy()` is a variant of `base::eval()`. The `eval_tidy()` help says: "`base::eval()` is sufficient for simple evaluation. Use `eval_tidy()` when you'd like to support expressions referring to the `.data` pronoun, or when you need to support quosure".
+ `rlang::parse_expr(string)` seems to do a similar job to `base::parse(text = string)`. They "transform text into R expressions...the first step in interpreting R code". `parse_quo()` also does this but creates a quosure.

Detail on tidyeval functions:

`sym()` and `expr()` both produce name objects but they are not the same. `sym()` is an evaluating function, while `expr()` is a quoting function. The argument to `sym()` must evaluate to a string. This string is converted to a name. `expr()` directly converts its argument to a name, without evaluating.

```{r sym}
x <- "mpg"
rlang::sym("word")
rlang::sym(x) 
rlang::expr(x)
rlang::expr("word") #this returns a character
```

Like `expr()`, `quo()` is a quoting function that does not evaluate its argument, but it returns a quosure, which is an expression together with and environment. `enquo()` takes an unquoted argument (a symbol), which it evaluates and returns a quosure with an expression formed from the contents of the object named by the symbol argument.

```{r enquo}
x <- "mpg"
rlang::enquo(x)
```

So to summarise:

```{r table, echo = FALSE}
dt <- tibble::tibble(
  func = c("quo", "expr", "enquo", "sym"),
  return = c("quosure", "name/symbol","quosure", "name/symbol"),
  `function type` = c("quoting", "quoting", "evaluating", "evaluating") 
) 
edwards::my_kable(dt)
```

I may be off in my definition of quoting/evaluating functions, but `enquo()` does evaluate its argument before quoting it. 

### Questions/surprises

`sym()` only works with strings so why does the following work?  

```{r q1}
x <- expr(mpg)
sym(x)
```

`x` has type symbol. Replacing `expr(mpg)` with `quo(mpg)` or `quo("mpg")` does error. since `sym()` cannot work with a quosure. Note that I have tried to use this expr/sym combo to solve the ggplot problem below but it is not a solution.

## Usage

Here are some examples of quote-unquoting with `select()`. The first 3 all select column `mpg`. The 4th select errors because it passes the string `"x"` to `select()`. 

```{r select1, eval=FALSE}
dt <- mtcars %>% as_tibble()
x <- "mpg"
select(dt, !!sym(x))
select(dt, !!expr(x))
select(dt, !!enquo(x))
select(dt, !!expr("x")) #error - col x does not exist
```

The next 4 are less clear. `sym()` and `enquo()` both evaluate `vs`, but `expr(vs)` does not. If there was no column called `vs` then `expr()` would look for other objects called `vs` and evaluate those. The third example errors because `enquo()` tries to evaluate its argument (same with `sym()`).  

```{r select2, eval = FALSE}
vs <- "mpg"
select(dt, !!sym(vs)) # mpg
select(dt, !!enquo(vs)) # mpg
select(dt, !!enquo(cyl)) # error - object `cyl` not found.
select(dt, !!expr(vs)) # selects vs
```

There isn't one standard for handling tidyselect inputs across the tidyverse. `select()` is very flexible - from its help page: *For convenience it also supports strings and character vectors. This is unlike other verbs where strings would be ambiguous*. Other functions may differ, notably `ggplot()` which takes unquoted names for variables, but not strings or integer positions. If these are supplied, e.g. `x = "name"`, then it will draw the plot but treat x as a categorical variable with fixed value `"name"`. 

Due to this, I have had issues using ggplot inside functions with dynamic naming. A simplified example is the following which takes the variables to plot as function arguments. It works with unquoted arguments but fails with strings, as might happen inside a shiny app, for example.

The following functions are simplified versions of what I am looking to do. The aim is to produce a scatter plot using named variables. The first, `myplot1()` works with unquoted arguments, but is incorrect with string arguments. Note that `{{` is equivalent to `!!enquo()`. The second, `myplot2()` works with string arguments but errors with unquoted arguments. 

```{r func1, eval = FALSE}
myplot1 <- function(dt, x, y) {
  gg <- ggplot(dt, aes({{x}}, {{y}})) + 
    geom_point()
  print(gg)
}
```

```{r func2, eval = FALSE}
myplot2 <- function(dt, x, y) {
  gg <- ggplot(dt, aes(!!sym(x), !!sym(y))) + 
    geom_point()
  print(gg)
}
```

Using `expr()` in place of `sym()` errors in both cases (`select()` would handle this but `ggplot()` is a much less flexible environment). Note ggplot2 only introduced tidyeval us in [version 3.0.0 blog post](https://www.tidyverse.org/blog/2018/07/ggplot2-tidy-evaluation/). That article says that variables are stored as quosures.

Other verbs like `filter()` work like ggplot, not `select()` and behaviour can be surprising at first. 

```{r filter, eval = F}
x <- "vs"
filter(mtcars, !!enquo(x) == 1) #empty df
filter(mtcars, !!sym(x) == 1) #works
```

What is happening is that `!!enquo(x)` (or equivalently `{{`) is evaluating to the string `"vs"` and `"vs" == 1` is false. Like `ggplot()`, `filter()` doesn't process strings as variable names due, I assume, to ambiguity. Because `sym()` creates a symbol this isn't evaluating to a string.

The following is a workaround for the ggplot issue. I use `select()` to handle the tidyselect input, extract the column names, then use these with `ggplot()`. This works fine but it seems crude way to do it. 

```{r func3, eval = FALSE}
myplot3 <- function(dt, x, y) {
  x <- names(select(dt, {{x}}))
  y <- names(select(dt, {{y}}))
  gg <- ggplot(dt, aes(!!sym(x), !!sym(y))) + 
    geom_point()
  print(gg)
}
```

Before looking at alternatives there is still the question of what is the appropriate way for this sort of function to work. Should I commit to only accepting unquoted inputs to avoid any ambiguity? Also, at some point I will need to consider how different tidyselect methods are affected by grouping.

The way `select()` internally processes tidyselection ([select github page](https://github.com/tidyverse/dplyr/blob/master/R/select.R)) is using the `eval_select()` function to return integer column location. The returned vector is named with the column names so, rather than using `select()`, I could create a function to just return the named location vector and work with that. Not necesarily simpler though.

```{r col_select}
col_select <- function(data, ...) {
  tidyselect::eval_select(expr(c(...)), data)
col_select(mtcars, 1, wt, "vs")
}
```

[Dynamic dots](https://rlang.r-lib.org/reference/dyn-dots.html) explains how `...` can be captured using `c()` or `list()`. Other functions used are `list2()`, `dots_list()`, `vars()`, and `call2()`.

[Filtering with string statements in dplyr](https://www.r-bloggers.com/2020/05/filtering-with-string-statements-in-dplyr/). I wanted to pass a string object to `dplyr::filter()` and this post gives a solution `filter(rlang::eval_tidy(rlang::parse_expr(selection_statement)))`. This can be done in base using `filter(eval(parse(text = selection_statement)))`. From the `eval_tidy()` I'd say `eval()` is fine here.

## NSE outside tidyverse functions

_Needs expanding_

The following works using base R:
```{r}
f <- function(nm) {
  ls <- list(a = 1, b = 2, c = 3)
  ls[[deparse(substitute(nm))]]
}
f(a)
```

I don't know what the rlang approach is. There is `rlang::expr_deparse()` <https://www.tidyverse.org/blog/2018/03/rlang-0.2.0/>.

[This blog post](https://www.r-bloggers.com/2019/08/no-visible-binding-for-global-variable/) talks about avoiding R CMD check notes when using NSE.

## Notes on Older Methods

The old NSE approach in dplyr was to have separate functions for SE e.g. `filter_()`. These underlay the NSE versions that were commonly used.

[Tidy evaluation in ggplot2](https://www.tidyverse.org/blog/2018/07/ggplot2-tidy-evaluation/). Tidyverse blog from mid-2018 when ggplot2 switched to using tidy eval operators and introduced `vars()`. 
The [lazyeval package](https://cran.r-project.org/web/packages/lazyeval/index.html): "This document outlines my [Hadley Wickham] previous approach to non-standard evaluation (NSE). You should avoid it unless you are working with an older version of dplyr or tidyr."

The following are some notes from the [lazyeval vignette](https://cran.r-project.org/web/packages/lazyeval/vignettes/lazyeval.html).

Common uses of NSE:

+ Labelling of plots and tables where the expression supplied to a function is used as a label rather than their value.
+ Non-standard scoping to look for objects in places other than the current environment e.g. `with()`, `subset()`, `transform()` start the search for objects in the current data frame or list.
+ Metaprogramming is a catch-all term that involves computing on the unevaluated code in some way.

In base R, an argument can be turned into a label using `deparse(substitute(x))`. Two problems are that `deparse()` may generate a vector of length > 1 and `substitue()` only looks one level up.
